---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

ComBat-refQL expects raw counts with genes in rows and samples in columns. Batch,
group, and covariate metadata must align with the sample columns.

```{r, eval=FALSE}
library(combatrefql)

fit <- combat_ref_ql(
  counts,
  batch = metadata$batch,
  group = metadata$condition
)

fit
summary(fit)
corrected <- fit@counts
```

To suppress fitting messages:

```{r, eval=FALSE}
fit <- combat_ref_ql(counts, batch = metadata$batch,
  group = metadata$condition, verbose = FALSE)
```

Set an explicit target with `reference = "batch_name"`; otherwise the method
selects a reference using biologically adjusted within-batch fits. Reference
counts remain unchanged.

`fit@gene_status` records adjusted, unsupported, and failed genes.
`fit@diagnostics` contains reference scores, raw and moderated batch contrasts,
dispersion multipliers, mapping events, and timings.

## Downstream analysis

For differential-expression inference, use the original counts and include
batch directly in the statistical design. This retains the observed count
observations and carries batch structure through the inferential model used by
edgeR, limma-voom, DESeq2, or an equivalent workflow.

Adjusted counts are useful for analyses that operate directly on an expression
matrix and benefit from a harmonized feature space, including PCA, clustering,
heatmaps, visualization, machine learning, and integrative analysis.

```{r, eval=FALSE}
log_cpm <- edgeR::cpm(fit@counts, log = TRUE, prior.count = 2)
pca <- prcomp(t(log_cpm), scale. = FALSE)
```

For predictive modelling, perform data-dependent preprocessing within the
training workflow and apply the learned procedure to held-out samples.

Fractional counts are audibly rounded by default. Use
`fractional_counts = "error"` to reject them. Set `verbose = FALSE` to suppress
fitting messages. `chunk_size` is an advanced transport-memory option that does
not change results.
