---
title: "Getting started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

ComBat-refQL expects raw counts with genes in rows and samples in columns. Batch,
group, and covariate metadata must align with the sample columns.

```{r, eval=FALSE}
library(combatrefql)

fit <- combat_ref_ql(
  counts,
  batch = metadata$batch,
  group = metadata$condition
)

fit
summary(fit)
corrected <- fit@counts
```

To suppress fitting messages:

```{r, eval=FALSE}
fit <- combat_ref_ql(counts, batch = metadata$batch,
  group = metadata$condition, verbose = FALSE)
```

Set an explicit target with `reference = "batch_name"`; otherwise the method
selects a reference using biologically adjusted within-batch fits. Reference
counts remain unchanged.

`fit@gene_status` records adjusted, unsupported, and failed genes.
`fit@diagnostics` contains reference scores, raw and moderated batch contrasts,
dispersion multipliers, mapping events, and timings.

Fractional counts are audibly rounded by default. Use
`fractional_counts = "error"` to reject them. Set `verbose = FALSE` to suppress
fitting messages. `chunk_size` is an advanced transport-memory option that does
not change results.
