---
title: "Input sanity and correction confidence"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Input sanity and correction confidence}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

## Input expectations

ComBat-refQL accepts raw bulk RNA-seq counts with genes in rows and samples in
columns. Counts must be finite, non-negative, and identified by unique gene and
sample names. Each sample belongs to exactly one discrete batch. `group` and
sample-by-covariate columns are optional preserved biology.

```{r, eval=FALSE}
fit <- combat_ref_ql(counts, batch) # zero-covariate workflow
fit <- combat_ref_ql(counts, batch, group, covariates)
```

The package errors on invalid counts or identifiers, metadata misalignment,
fewer than two batches, invalid references, normalized/transformed expression
when confidently recognizable, exact rank deficiency, and zero residual degrees
of freedom. Constant terms and unused factor levels are recorded in
`fit@diagnostics$input`; constant terms are omitted because they preserve no
sample-varying information. Library-size outliers are flagged relative to the
median (below one quarter or above four times it), not by a universal count
cutoff. A batch-size ratio of at least four is reported as highly unequal.

## Batch replication

Singleton source batches are accepted when the global design remains
identifiable. Their source-batch dispersion uses the existing shared/reference
fallback and is not independently estimable. Singletons are never eligible for
automatic reference selection and their corrections are labelled low
confidence. Two-sample sources are accepted but labelled limited-information;
their confidence cannot be high. Larger batches are assessed with the existing
weighted residual-information diagnostic rather than a sample-count rule.

An explicitly selected singleton reference generally has zero independently
estimable reference dispersion information and produces an actionable error;
choose a replicated reference.

## Biology and batch

Exact confounding is decided only by matrix rank. The additional entanglement
score is descriptive: for each non-batch design column, ComBat-refQL reports the
fraction of its centered sum of squares explained by the batch-design subspace,
then takes the maximum across columns belonging to a term. Scores are bounded in
`[0,1]`. A maximum at or above `0.8` produces a warning; this conservative value
is a diagnostic category, not a statistical boundary.

```{r, eval=FALSE}
# balanced
batch <- factor(rep(c("A", "B"), each = 6))
group <- factor(rep(c("control", "treated"), 6))
combat_ref_ql(counts, batch, group)

# partial association: proceeds and is recorded
group <- factor(c("control", "control", "treated", "treated", "control", "treated",
                  "control", "treated", "treated", "control", "treated", "control"))
combat_ref_ql(counts, batch, group)

# strong but not exact association: proceeds with a warning
batch <- factor(rep(c("A", "B"), each = 24))
group <- factor(c(rep("control", 23), "treated", "control", rep("treated", 23)))
combat_ref_ql(counts48, batch, group)

# exact aliasing: informative combatrefql_design_error
combat_ref_ql(counts, batch, group = batch)
```

With exact confounding the package cannot determine which inseparable component
is biology and which is batch. It therefore does not silently drop either term.
Locally non-estimable biological terms used only during automatic reference
scoring are recorded in the reference table together with fallbacks, scores,
runner-up, and score margin.

## Correction confidence

Confidence is reported for each `gene x source_batch -> reference_batch`
correction. It reuses the fitted QL standard error `s`, weighted effective
information `I`, dispersion estimability, mapping validity, and source batch
size. No model is refitted.

The exact diagnostic reliability/support index is

```
min(1 / (1 + s), I / (1 + I), batch_cap, dispersion_cap)
```

where `batch_cap` is `0.25`, `0.50`, or `1` for source sizes 1, 2, or at least
3, and `dispersion_cap` is `1` when independently estimable and `0.25` when the
existing fallback is used. A mapping failure sets the score to zero and the
label to `failed`. Otherwise labels are `high` for scores at least `0.75`,
`moderate` for scores at least `0.50`, and `low` below `0.50`.

The score is a deterministic reliability/support index, not a posterior
probability, p-value, or test of whether a batch effect exists. Raw effects,
standard and adaptive EB weights, and posterior effects remain visible in the
same row. Adaptive weight describes how the estimate combines gene evidence
and the fitted prior; it is deliberately not treated as evidence that the
correction is reliable, so a small estimated batch effect is not automatically
penalized.

`fit@gene_status` conservatively uses the minimum score and worst label across
all source corrections for each gene. Full detail remains available:

```{r, eval=FALSE}
fit@gene_status
fit@diagnostics$correction_confidence
fit@diagnostics$batch_quality
fit@diagnostics$input
fit@diagnostics$design_entanglement
summary(fit)
```

These diagnostics expose limited support; they do not remove the fundamental
non-identifiability of exactly confounded designs or create replication where
none exists.
